### 2020-04-08 thesis/content/chapter-ltd.tex
started with real-world eval section + various improvements

	The resulting feature matrix $Z^{(t,2+K)}$ then satisfies all three invariants which completes the induction.

@@

To complete our analysis of the expressive power of 2-\acs{wl}-\acsp{gnn} we now show that they are not just as powerful as 1-\acs{wl} but in fact more powerful than 1-\acs{wl}.
\begin{prop}\label{cor:ltd:wl2-gnn-regular}
	There are $d$-regular graphs $G$ and $H$ of size $n$ which can be distinguished by 2-\acs{wl}-\acsp{gnn}.
\end{prop}
\begin{proof}
	The proposition follows if we choose the six-cycle graph for $G$ and the two three-cycles graph for $H$, both of which should be well-known at this point.
	Let $h_2 = \mathit{Pool} \circ S$ be a 2-\acs{wl}-\acs{gnn} with the neighborhood radius $r = 1$ which consists of a single 2-\acs{wl} convolution layer $S: \mathbb{R}^{* \times 2} \to \mathbb{R}^{* \times 1}$ and the pooling layer $\mathit{Pool} = \min$.
	In accordance with \cref{defn:ltd:wl2-conv-init} we set the initial feature vectors of the vertices $v_i$ of $G$ and $H$ to $Z^{(0)}[e_{ii}] \coloneqq (1, 0)$ and the initial feature vectors of their edges $e_{ij}$ to $Z^{(0)}[e_{ij}] \coloneqq (0, 1)$.
	Let the weight matrices of $S$ be $W_{\mathrm{L}} \coloneqq \mathbf{0}$ and $W_{\mathrm{F}} = W_{\Gamma} \coloneqq \begin{pmatrix} 1\\1 \end{pmatrix}$.
	For simplicity we choose the identity activation functions $\sigma = \sigma_{\Gamma} = \mathrm{id}$.

	By \cref{defn:ltd:wl2-conv-step} all self-loops $e_{ii}$ of $G^1$ and $H^1$ have the three neighbors $\{ \textcolor{t_blue}{\ldblbrace e_{ii}, e_{ii} \rdblbrace},\allowbreak \textcolor{t_red}{\ldblbrace e_{ij}, e_{ji} \rdblbrace},\allowbreak \textcolor{t_darkgreen}{\ldblbrace e_{il}, e_{li} \rdblbrace} \}$, i.e.\ the length-two walk along $e_{ii}$ itself and the length-two walks to and from the two neighboring vertices $\Gamma(v_i) = \{ v_j, v_l \}$. % chktex 25
	Therefore the convolved feature vector of all self-loops are $Z^{(1)}[e_{ii}] = (1) \odot (\textcolor{t_blue}{(1 + 1)} + \textcolor{t_red}{(1 + 1)} + \textcolor{t_darkgreen}{(1 + 1)}) = 6$. % chktex 25
	However for the non-self-loops of $G^1$ and $H^1$, i.e.\ the edges of $G$ and $H$, we get differing convolved feature vectors.
	The 2-\acs{wl} neighbors of $e_{ij} \in \mathcal{E}_G$ are $\{ \ldblbrace e_{ii},  e_{ij} \rdblbrace,\allowbreak \ldblbrace e_{ij}, e_{jj} \rdblbrace \}$.
	The 2-\acs{wl} neighbors of $e_{ij}' \in \mathcal{E}_H$ are $\{ \ldblbrace e_{ii}',  e_{ij}' \rdblbrace,\allowbreak \ldblbrace e_{ij}',  e_{jj}' \rdblbrace,\allowbreak \ldblbrace e_{il}',  e_{lj}' \rdblbrace \}$ where $v_l' \in \mathcal{V}_H$ is the common neighbor of $v_i'$ and $v_j'$.
	The different neighborhood sizes of the edges of $G$ and $H$ imply that $\forall e_{ij} \in \mathcal{E}_G: Z^{(1)}[e_{ij}] = 4$ while $\forall e_{ij}' \in \mathcal{E}_H: Z^{(1)}[e_{ij}] = 6$.
	Thus $h_2(G) = \min\{ 4, 6 \} \neq \min\{ 6, 6 \} = h_2(H)$ which concludes the proof. % chktex 21
\end{proof}
\begin{cor}\label{cor:ltd:wl2-gnn-more-wl1-power}
	The discriminative power of 2-\acs{wl}-\acsp{gnn} is strictly greater than that of the 1-\acs{wl} algorithm.
\end{cor}
\begin{proof}
	The corollary directly follows from \cref{cor:ltd:wl2-gnn-wl1-power} and \cref{cor:ltd:wl2-gnn-regular} since 1-\acs{wl} cannot distinguish regular graphs~\cite[cor.~1.8.5]{Immerman1990}.
\end{proof}

@@

Using the $\mathit{gather}$ and $\mathit{scatter}_{\Sigma}$ operator, the 2-\acs{wl} convolution operator from \cref{defn:ltd:wl2-conv-step} can be computed via the following algorithm:
\begin{algorithm}[H]
	\caption{Parallel Implementation of a 2-\acs{wl} Convolution Layer $S^{(t)}$}\label{algo:ltd:wl2-conv}
	\begin{algorithmic}[1]
		\Function{$S^{(t)}$}{$Z^{(t-1)} \in \mathbb{R}^{m \times d^{(t-1)}}, R \in {[m]}^{\gamma \times 3}$}
			\State{$Z_{\mathrm{L}} \coloneqq Z^{(t-1)} W_{\mathrm{L}}^{(t)}$}
			\Comment{Matrix multiply: $\mathbb{R}^{m \times d^{(t-1)}} \to \mathbb{R}^{m \times d^{(t)}}$}
			\State{$Z_{\mathrm{F}} \coloneqq Z^{(t-1)} W_{\mathrm{F}}^{(t)}$}
			\State{$Z_{\Gamma} \coloneqq Z^{(t-1)} W_{\Gamma}^{(t)}$}
			\State{$X_{\Gamma, 1} \coloneqq \mathit{gather}(Z_{\Gamma}, R_{\Gamma, 1})$}
			\Comment{Gather: $\mathbb{R}^{m \times d^{(t)}} \times {[m]}^{\gamma} \to \mathbb{R}^{\gamma \times d^{(t)}}$}
			\State{$X_{\Gamma, 2} \coloneqq \mathit{gather}(Z_{\Gamma}, R_{\Gamma, 2})$}
			\State{$X_{\Gamma} \coloneqq \sigma_{\Gamma}\left(X_{\Gamma, 1} + X_{\Gamma, 2}\right)$}
			\Comment{Element-wise operations}
			\State{$Z_{\Sigma\Gamma} \coloneqq \mathit{scatter}_{\Sigma}(X_{\Gamma}, R_{\mathrm{L}})$}
			\Comment{Scatter: $\mathbb{R}^{\gamma \times d^{(t)}} \times {[m]}^{\gamma} \to \mathbb{R}^{m \times d^{(t)}}$}
			\State{$Z^{(t)} \coloneqq \sigma\left(Z_{\mathrm{L}} + Z_{\mathrm{F}} \odot Z_{\Sigma\Gamma} \right)$}
			\Comment{Element-wise operations}
			\State{\Return{$Z^{(t)}$}}
		\EndFunction{}
	\end{algorithmic}
\end{algorithm}
All operations in this implementation scale well on parallel computing hardware, they are differentiable and are supported by common \ac{ml} libraries like \citetitle{TF}~\cite{Abadi2015}\cite{TF} and \citetitle{PyT}~\cite{Paszke2019}\cite{PyT}.
2-\acs{wl} convolutions can therefore be easily integrated into existing \ac{ml} tooling and be optimized via well-known gradient-based methods like \textit{Adam}~\cite{Kingma2015}.
This concludes our description of 2-\acs{wl}-\acsp{gnn}.

>>>>

	The resulting feature matrix $Z^{(t,2+K)}$ then satisfies all three invariants, which completes the induction.

@@

To complete our analysis of the expressive power of 2-\acs{wl}-\acsp{gnn} we now show that they are not just as powerful as 1-\acs{wl} but in fact more powerful than 1-\acs{wl}.
\begin{prop}\label{cor:ltd:wl2-gnn-regular}
	There are $d$-regular graphs $G$ and $H$ of size $n$ which can be distinguished by 2-\acs{wl}-\acsp{gnn}.
\end{prop}
\begin{proof}
	The proposition follows if we choose the six-cycle graph for $G$ and the two three-cycles graph for $H$, both of which should be familiar at this point.
	Let $h_2 = \mathit{Pool} \circ S$ be a 2-\acs{wl}-\acs{gnn} with the neighborhood radius $r = 1$ which consists of a single 2-\acs{wl} convolution layer $S: \mathbb{R}^{* \times 2} \to \mathbb{R}^{* \times 1}$ and the pooling layer $\mathit{Pool} = \min$.
	In accordance with \cref{defn:ltd:wl2-conv-init} we set the initial feature vectors of the vertices $v_i$ of $G$ and $H$ to $Z^{(0)}[e_{ii}] \coloneqq (1, 0)$ and the initial feature vectors of their edges $e_{ij}$ to $Z^{(0)}[e_{ij}] \coloneqq (0, 1)$.
	Let the weight matrices of $S$ be $W_{\mathrm{L}} \coloneqq \mathbf{0}$ and $W_{\mathrm{F}} = W_{\Gamma} \coloneqq \begin{pmatrix} 1\\1 \end{pmatrix}$.
	For simplicity we choose the identity activation functions $\sigma = \sigma_{\Gamma} = \mathrm{id}$.

	By \cref{defn:ltd:wl2-conv-step}, all self-loops $e_{ii}$ of $G^1$ and $H^1$ have the three neighbors $\{ \textcolor{t_blue}{\ldblbrace e_{ii}, e_{ii} \rdblbrace},\allowbreak \textcolor{t_red}{\ldblbrace e_{ij}, e_{ji} \rdblbrace},\allowbreak \textcolor{t_darkgreen}{\ldblbrace e_{il}, e_{li} \rdblbrace} \}$, i.e.\ the length-two walk along $e_{ii}$ itself and the length-two walks to and from the two neighboring vertices $\Gamma(v_i) = \{ v_j, v_l \}$. % chktex 25
	Therefore the convolved feature vector of all self-loops are $Z^{(1)}[e_{ii}] = (1) \odot (\textcolor{t_blue}{(1 + 1)} + \textcolor{t_red}{(1 + 1)} + \textcolor{t_darkgreen}{(1 + 1)}) = 6$. % chktex 25
	However for the non-self-loops of $G^1$ and $H^1$, i.e.\ the edges of $G$ and $H$, we get differing convolved feature vectors.
	The 2-\acs{wl} neighbors of $e_{ij} \in \mathcal{E}_G$ are $\{ \ldblbrace e_{ii},  e_{ij} \rdblbrace,\allowbreak \ldblbrace e_{ij}, e_{jj} \rdblbrace \}$.
	The 2-\acs{wl} neighbors of $e_{ij}' \in \mathcal{E}_H$ are $\{ \ldblbrace e_{ii}',  e_{ij}' \rdblbrace,\allowbreak \ldblbrace e_{ij}',  e_{jj}' \rdblbrace,\allowbreak \ldblbrace e_{il}',  e_{lj}' \rdblbrace \}$ where $v_l' \in \mathcal{V}_H$ is the common neighbor of $v_i'$ and $v_j'$.
	The different neighborhood sizes of the edges of $G$ and $H$ imply that $\forall e_{ij} \in \mathcal{E}_G: Z^{(1)}[e_{ij}] = 4$ while $\forall e_{ij}' \in \mathcal{E}_H: Z^{(1)}[e_{ij}] = 6$.
	Thus $h_2(G) = \min\{ 4, 6 \} \neq \min\{ 6, 6 \} = h_2(H)$, which concludes the proof. % chktex 21
\end{proof}
\begin{cor}\label{cor:ltd:wl2-gnn-more-wl1-power}
	The discriminative power of 2-\acs{wl}-\acsp{gnn} is strictly greater than that of the 1-\acs{wl} algorithm.
\end{cor}
\begin{proof}
	The corollary directly follows from \cref{cor:ltd:wl2-gnn-wl1-power} and \cref{cor:ltd:wl2-gnn-regular} since 1-\acs{wl} cannot distinguish regular graphs~\cite[cor.~1.8.5]{Immerman1990}.
\end{proof}

@@

Using the $\mathit{gather}$ and $\mathit{scatter}_{\Sigma}$ operators, the 2-\acs{wl} convolution operator from \cref{defn:ltd:wl2-conv-step} can be computed via the following algorithm:
\begin{algorithm}[H]
	\caption{Parallel Implementation of a 2-\acs{wl} Convolution Layer $S^{(t)}$}\label{algo:ltd:wl2-conv}
	\begin{algorithmic}[1]
		\Function{$S^{(t)}$}{$Z^{(t-1)} \in \mathbb{R}^{m \times d^{(t-1)}}, R \in {[m]}^{\gamma \times 3}$}
			\State{$Z_{\mathrm{L}} \coloneqq Z^{(t-1)} W_{\mathrm{L}}^{(t)}$}
			\Comment{Matrix multiply: $\mathbb{R}^{m \times d^{(t-1)}} \to \mathbb{R}^{m \times d^{(t)}}$}
			\State{$Z_{\mathrm{F}} \coloneqq Z^{(t-1)} W_{\mathrm{F}}^{(t)}$}
			\State{$Z_{\Gamma} \coloneqq Z^{(t-1)} W_{\Gamma}^{(t)}$}
			\State{$X_{\Gamma, 1} \coloneqq \mathit{gather}(Z_{\Gamma}, R_{\Gamma, 1})$}
			\Comment{Gather: $\mathbb{R}^{m \times d^{(t)}} \times {[m]}^{\gamma} \to \mathbb{R}^{\gamma \times d^{(t)}}$}
			\State{$X_{\Gamma, 2} \coloneqq \mathit{gather}(Z_{\Gamma}, R_{\Gamma, 2})$}
			\State{$X_{\Gamma} \coloneqq \sigma_{\Gamma}\left(X_{\Gamma, 1} + X_{\Gamma, 2}\right)$}
			\Comment{Element-wise operations}
			\State{$Z_{\Sigma\Gamma} \coloneqq \mathit{scatter}_{\Sigma}(X_{\Gamma}, R_{\mathrm{L}})$}
			\Comment{Scatter: $\mathbb{R}^{\gamma \times d^{(t)}} \times {[m]}^{\gamma} \to \mathbb{R}^{m \times d^{(t)}}$}
			\State{$Z^{(t)} \coloneqq \sigma\left(Z_{\mathrm{L}} + Z_{\mathrm{F}} \odot Z_{\Sigma\Gamma} \right)$}
			\Comment{Element-wise operations}
			\State{\Return{$Z^{(t)}$}}
		\EndFunction{}
	\end{algorithmic}
\end{algorithm}
All operations in this implementation scale well on parallel computing hardware, they are differentiable and are supported by common \ac{ml} libraries like \citetitle{TF}~\cite{Abadi2015}\cite{TF} and \citetitle{PyT}~\cite{Paszke2019}\cite{PyT}.
2-\acs{wl} convolutions can therefore be easily integrated into existing \ac{ml} tooling and be optimized via well-known gradient-based methods like \textit{Adam}~\cite{Kingma2015}.
This concludes our description of 2-\acs{wl}-\acsp{gnn}.
